Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-02-20 paper.qmd
worked on uncertainty vis and grammar of graphics section

## Uncertainty Visualisation and the Grammar of Graphics
The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty.

>>>>

## Uncertainty Visualisation and the Grammar of Graphics
The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty. This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments. How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies? The answer circles back to the definition problem we discussed earlier.

There are two ways to think of uncertainty in graphics:
1) A parameter is defined and graphed, and uncertainty is depicted as additional information in that graphic. Therefore uncertainty visualisation is a layering a "normal" visualisation with some aesthetic changes that incorperate uncertainty.
2) Uncertainty is defined (using a PDF, confidence interval, etc) and graphed using whatever method is required.
The difference between these two conceptualisations stems from the two different ways to percieve the uncertainty visualisation problems. Method (1) requires thinking of uncertainty as a dimensionality problem, while (2) sees uncertainty as something that is defined and visualised on its own. For example, someone who percieves of uncertainty visualisation as (1) would see a pie chart as a normal graphic, and a pie chart with blurred boundaries as an uncertainty visualisation, while someone who uses method (2) would see a pie chart that depicts probabilities of an event as an uncertainty graphic and blurred boundaries would be pointless.

These two differing philosophies diverge in *where* in the visualisation pipeline they believe uncertainty to be defined. Those who are using method (1) believe uncertainty is defined at the geometry or aesthetics stages shown in @fig-pipeline, while (2) believes uncertainty is defined at the statistics stage. Unfortunately there is no single person or experiment that is consistent in any one belief. Papers will simultanously discuss uncertainty as though a visualisation of variance is an uncertainty visualisation but also assume a bar char of variances is *not* an uncertainty visualisation. This difficulty and lack of consistency comes from the desire to discuss "uncertainty visualisation" as though it is a single term and not the combination of "uncertainty" and "visualisation". If we split the terms and look at the quantification and visualisation of uncertainty separately, the results of many uncertainty visualisation papers become self evident.

![The grammar of graphics data analysis pipeline](grammarofgraphicspipeline.png){#fig-pipeline}

Regardless of how you see uncertainty visualisations, there are two things that must be done in order to depict uncertainty in a graphic.
1) Mathematically estimate uncertainty
2) Visualise uncertainty
There are numerous ways to depict uncertainty mathematically. Examples of mathematically defined objects that are used to express uncertainty inlude: confidence intervals, sampling distributions, bias, variance, prediction error, value ranges, quantiles etc. These things can all be specificed, calculated, and set as the ground truth in a visualisation experiment. There are also numerous ways to depict uncertainty visually. The computed values can be mapped to a position, length, colour, area, etc. This process, of quantifying what you are visualising and then attributing it to a visual attribute is not unique to uncertainty visualisation. In this sense, uncertainty is exactly the same as any other variable, and if in this sense, uncertainty visualisation is exactly the same as normal visualisation.

There are also numerous ways to something we have mathematically depicted.
